---
title: "Final Project"
author: "Jennifer Frediani"
date: "3/4/2020"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
```{r}
#load data from SPSS
library(ggplot2)
library(dplyr)
library(haven)
hints4cycle2_09062017_public_edited <- read_sav("~/CPTR-NRSG 741/HINTS4-Cycle2-SPSS/hints4cycle2_09062017_public_edited.sav")
hints4cycle2 <- hints4cycle2_09062017_public_edited
```
These data are from the Health Information National Trends Survey (HINTS)data from the National Cancer Institute.  I chose HINTS 4 Cycle 2 as a starting point for these analyses because it had questions on BMI, cancer history, questions pertaining to screening history and information on diet and exercise. I am interested in exploring relationships between healthy lifestyle behaviors and incidence of prostate cancer in people with and without a family history. 

This loaded dataset was simplified in SPSS before importing to R.

```{r}
# Now to explore

class(hints4cycle2)
dim(hints4cycle2)
```

There are 3630 observations in this dataset. For now we will limit the dataset to only those that have or had prostate cancer.

```{r}
hints4cycle2 %>%
 filter(CaProstate == 1) %>%
 nrow()
```
There are 60 participants that have or had prostate cancer. We'll now look at of those 60 who has a family history of cancer.

```{r}
hints4cycle2 %>%
 filter(CaProstate == 1) %>%
  filter(FamilyEverHadCancer == 1) %>%
 nrow()
```
Well, the majority of those who have/had prostate cancer also had a family history of cancer. So let's look at healthy behaviors in those with prostate cancer.

```{r}
#fruit intake  (3 = at least 2 cups)
hints4cycle2 %>%
 filter(CaProstate == 1) %>%
  filter(Fruit == 3) %>%
  nrow ()
```
```{r}
#vegetable intake (4 = at least 3 cups)
hints4cycle2 %>%
 filter(CaProstate == 1) %>%
  filter(Vegetables == 4) %>%
  nrow ()
```
So we have poor fruit and vegetable intake, not suprising, how are they doing with exercise...

```{r}
# First we have to calculate whether they meet the PA guidelines 
#meetExGuide <- 0
 # for(PersonID in hints4cycle2) {
  #  meetExGuide <- hints4cycle2$TimesModerateExercise * hints4cycle2$HowLongModerateExerciseHr
  #}  
# print(meetExGuide)

#exercise (moderate by hour)
hints4cycle2 %>%
 filter(CaProstate == 1) %>%
  filter(HowLongModerateExerciseHr > 2) %>%
  nrow ()
```
Not many...

Now let's look at those with a family history of cancer but have not had cancer yet.

```{r}
hints4cycle2 %>%
  filter(FamilyEverHadCancer == 1) %>%
  filter(EverHadCancer == 2)%>%
 nrow()
```
```{r}
#fruit intake  (3 = at least 2 cups)
hints4cycle2 %>%
 filter(FamilyEverHadCancer == 1) %>%
  filter(EverHadCancer == 2)%>%
  filter(Fruit == 3) %>%
  nrow ()
```
```{r}
#vegetable intake  (4 = at least 3 cups)
hints4cycle2 %>%
 filter(FamilyEverHadCancer == 1) %>%
  filter(EverHadCancer == 2)%>%
  filter(Vegetables == 4) %>%
  nrow ()
```
```{r}
#exercise
hints4cycle2 %>%
 filter(FamilyEverHadCancer == 1) %>%
  filter(EverHadCancer == 2)%>%
   filter(HowLongModerateExerciseHr > 2) %>%
  nrow ()
```

```{r}
hints4cycle2$EverHadCancer <- as.numeric(hints4cycle2$EverHadCancer)
ggplot(hints4cycle2, aes(x=Fruit, y=Vegetables, color=EverHadCancer)) 
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
